---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo=TRUE
)
library(rsimpop)
```
---
date: `r format(Sys.Date(), "%d/%m/%Y")`
---


# rsimpop

<!-- badges: start -->
<!-- badges: end -->


## Installation

You can install rsimpop like so:

``` r
devtools::install_github("NickWilliamsSanger/rsimpop")
```

## rsimpop
This package facilitates the simultaneous simulation of multiple cellular compartments each with their own target population size and potentially also sub-compartments with differential fitness (driver compartments).

## Simulate from Zygote for 1 year and subsample tree
```{r zygote,results=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}

##Initialise with seed (R and rsimpop separately)
SEED=37774323
initSimPop(SEED,bForce = TRUE)
##Setup a single compartment with a target pop of 50K
cfg=getDefaultConfig(target_pop_size  = 5e4,ndriver = 1,basefit = 0.2,rate = 0.1)
print(cfg)
##Simulate for 2years..
sp=sim_pop(NULL,params=list(n_sim_days=365*2,b_stop_at_pop_size=1),cfg=cfg)
##Look at the population size trajectory
plot(sp)
##Subsample tree
sampledtree1=get_subsampled_tree(sp,100)
print(sampledtree1)
plot_tree(sampledtree1,cex.label = 0.5)
title("Sampled Zygote Tree: Division Tree")
```

Notice how the sampled tree has 101 tips rather than the specified 100.  This is because the simulator now always maintains an inactive outgroup (here s1). A group is rendered inactive by specifying a negative "rate" in the cfg$compartment dataframe.  The tree branch lengths are now given in terms of the number of self renewal divisions.  This allows the user to flexibly apply their own mutation acquisition model:

```{r mutacq}
get_elapsed_time_tree
sampledtree1m=get_elapsed_time_tree(sampledtree1,mutrateperdivision=1,backgroundrate=15/365)
plot_tree(sampledtree1m,cex.label = 0.5);title("Sampled Zygote Tree: Mutation Tree")
```

Actually this illustrates a potential problem with the outgroup sample still having a acquired mutations because it has a finite duration (0 to 365 days)..

The changes between compartments is specified in a separate data.frame, tree$events , that is maintained and updated by the simulator.   

```{r events, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
t1=plot_tree(sampledtree1m,cex.label = 0.5);title("Sampled Zygote Tree: Mutation Tree")
node_labels(t1,cex=0.5)
print(sampledtree1m$events)
```

Notice how the events dataframe specifies the compartment for the outgroup and the rest of the tree.

We can introduce another cell compartment as follows:

```{r newcompartment,message=TRUE}
cfg=sampledtree1$cfg
cfg=addCellCompartment(cfg,population = 5e4,rate=1/50,ndriver=1,descr="MyTissue",basefit = 0.3)
cfg$compartment$rate[2]=1/120  ## change the rate of compartment 1
sampledtree1a=addDifferentiationEvents(sampledtree1,cfg,2,nEvent=10)
print(sampledtree1a$events)
```

Each branch carries its final compartment membership in the "state" vector.
```{r plotcompartment, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
sampledtree1a$color = c("grey","black","red")[sampledtree1a$state+1]
plot_tree(sampledtree1a,cex.label = 0.5);title("Highlights branches with compartment changes")
```
The above plot does not capture the situation when compartment changes take place mid-branch - so alternatively we can better visualise the situation using the built in function *plot_tree_events*
```{r plotc2, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
plot_tree_events(sampledtree1a)
```


We've already updated the config with the target population sizes and division rates, so we're ready to simulate:
```{r simagain,message=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
sp2=sim_pop(sampledtree1a,params=list(n_sim_days=365*10),cfg=sampledtree1a$cfg)
sp2=combine_simpops(sp,sp2)
plot(sp2)
sampledtree2=get_subsampled_tree(sp2,100)
plot_tree_events(sampledtree2,cex.label = 0.5)
```


## Selection based simulation
Here we are interested in the simple situation of one cellular compartment with multiple sub-compartments.

```{r selection,error=TRUE,echo=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
run_selection_sim
selsim=run_selection_sim(0.05,1/(2*190),target_pop_size = 5e4,nyears = 50,fitness=0.3)
print(selsim$cfg$info)
```

Plot a sampled tree
```{r seltree, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
seltree100=get_subsampled_tree(selsim,100)
print(seltree100$cfg$info)
plot_tree_events(seltree100,cex.label = 0);title("Selection Based Tree: Branch Length=#Self Renewal Divisions")
seltree100rt=get_elapsed_time_tree(seltree100)
tree=plot_tree_events(seltree100rt,cex.label = 0);title("Selection Based Tree: Branch Length=#Real Time")
mp=5
seltree100m=get_elapsed_time_tree(seltree100,mutrateperdivision=mp,backgroundrate=(20-(365/190)*mp)/365)
plot_tree_events(seltree100m,cex.label = 0.5);title("Selection Based Tree: Branch Length=#Mutations")
seltree100m2=get_elapsed_time_tree(seltree100,mutrateperdivision=20*(190/365),backgroundrate=0)
plot_tree_events(seltree100m2,cex.label = 0.5);title("Selection Based Tree: Branch Length=#Mutations v2")
```

## Transient selection

```{r selectiont,error=TRUE,echo=TRUE, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
run_transient_selection
tselsim=run_transient_selection(0.05,1/(2*190),target_pop_size = 5e4,nyears_driver_acquisition=15,
                                  nyears_transient_end=30,
                                  nyears=50,
                                  fitness=0.5)
tseltree200=get_subsampled_tree(seltree,200)
plot_tree_events(get_elapsed_time_tree(tseltree200),cex.label=0)
```


## Neutral simulation with a trajectory

Create a trajectory dataframe with 3 columns (ts,target_pop_size,division_rate) and simulate using the run_neutral_trajectory wrapper function. Note that timestamps and rates are expressed in units of days and expected divisions per day respectively.

```{r neutwithtraj, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
trajectory=data.frame(ts=365*(1:80),target_pop_size=5e4+100*(1:80),division_rate=1/(2*190))
trajectory$target_pop_size[5:10]=2*trajectory$target_pop_size[5:10]
trajectory$target_pop_size[11:15]=0.2*trajectory$target_pop_size[11:15]
print(head(trajectory))
sp=run_neutral_trajectory(NULL,0.5,trajectory)
plot(sp,xlim=c(0,100))
lines(trajectory$ts/365,trajectory$target_pop_size,col="red")
legend("topright",c("Target","Actual"),col=c("red","black"),lwd=1)
fulltree=get_tree_from_simpop(sp)
st=get_subsampled_tree(fulltree,100)
plot_tree(get_elapsed_time_tree(st),cex.label = 0)
```


## Multiple drivers

Multiple drivers can be generated at a specified rate so the waiting time between events is exponentially distributed.

Firstly the user need to create a function that draws a selection coefficient from a distibution. The simulator isn't optimised to maintain 100s of variants at once - so it is suggested that a minumum selective coefficient be specified (say 0.05) and driver incidence made correspondingly rarer.
```{r}
##Function to generate exponential distribution based fitness
require("truncdist")
genExpFitness=function(fitness_threshold,rate){
  function() rtrunc(n=1,a=fitness_threshold, b=Inf,"exp",rate=rate)
}
fitnessExpFn=genExpFitness(fitness_threshold=0.08,rate=40)
hist(sapply(1:100000,function(i) exp(fitnessExpFn())-1),breaks=seq(0,100,0.01),xlim=c(0,1),xlab="Selective Coefficient Per Year",main="Sampled Selective Cofficent Distribution")
```


Now run the sim:

```{r,message=FALSE,results=FALSE}
dps=run_driver_process_sim(0.1,1/(2*190),target_pop_size = 1e5,nyears = 80,fitness=fitnessExpFn,drivers_per_year = 1)
```

Look at the final per driver counts

```{r}
print(dps$cfg$info %>% filter(population>0))
```

Plot an example sampled tree

```{r, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
dpst=get_subsampled_tree(dps,200)
dpst=get_elapsed_time_tree(dpst)
plot_tree_events(dpst)
dpst=get_elapsed_time_tree(dpst,mutrateperdivision = 1,backgroundrate = 19/365)
plot_tree_events(dpst,fmode=1)
```

Continue simulating the same individual until the age of 90
```{r}
dps90=continue_driver_process_sim(dps,90,fitnessGen = fitnessExpFn)
```
Note the driver ids are reused once they become extinct so there is no guarantee that they are preserved between runs.
```{r, dev.args=list(pointsize=8),fig.height=5,fig.width=10}
dpst=get_subsampled_tree(dps90,200)
dpst=get_elapsed_time_tree(dpst)
plot_tree_events(dpst)
```



